Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-01-08 paper/paper.tex
pass on step 2

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,anonymous]{acmart}

@@

\noindent
\textbf{Differences.}~Note that there are differences between the
options produced in the first step of \tname{} and the options from
the golden rule, as per Figure~\ref{fig:synflood-example}. The option
\CodeIn{flags} does not mention parameters \CodeIn{1 2}, there is an
extra option \CodeIn{window}, and there are differences in the option
\CodeIn{threshold}. Considering the option \CodeIn{flags}, the first
parameter of that option indicates what flags in the packet must be
set while the second parameter indicates non-mandatory flags (\ie{},
``don't cares''). The option \CodeIn{flags: S} indicates that the flag
``S'' must be set and all flags not mentioned (``1'' and ``2'', in our
case) must be unset for the \nids\ to capture the
traffic. Consequently, that choice of option alone could result in
false negatives and false positives. An attack that satisfies all
other conditions in a rule and have flags ``1'' or ``2'' set would be
captured by the original rule but not the inferred rule---a case of
false negative. Likewise, an attack that satisfies all conditions and
have flags ``1'' and ``2'' unset would be captured by the inferred
rule but not the original rule---a case of false positive. The set of
positive examples is to blame in this case. To mitigate this issue,
the input messages should have covered cases with options ``1'' or
``2'' set. Considering option \CodeIn{windows: 64}, \tname{} is able
to generate rules without that option, as discussed above. Considering
option \CodeIn{threshold}, the values used in the golden rule are
domain-specific. \tname{} uses default values for these options. We
conjecture that the system administrator would be able to configure
this option.

@@

@@

\begin{figure*}[t!]
\centering
\includegraphics[trim=0 350 100 0,clip,width=0.65\textwidth]{figs/nids-workflow}
\caption{\tname\ workflow.}
\label{fig:overview}
\end{figure*}

@@

\Mar{-------------------------- estou aqui}

The rule reported in the first step satisfies the criterion to only
capture positive traffic. Unfortunately, the rule produced in the
first step is overconstrained---it is unable to capture different
manifestations of the same attack as the rule matches all options that
\tname\ found for the input message.

@@

\tname{} leverages public datasets\cite{tcpreplay} of negative traffic
to guide the search for more general rules, hence addressing the
overffiting issue. The goal of this step is to create rules that
captures traffic similar to the input malicious traffic. To achieve
that goal, \tname{} \emph{minimizes} the rules produced in the
previous step, preserving the invariant of \emph{only} capturing the
positive malicious traffic. A rule is minimal if discarding any option
results in capturing negative (\ie{}, benign) traffic.  The
minimization procedure we propose in this step builds on the
observation that all options from the rule produced in the previous
step are satisifed by the positive traffic. Consequently, any subset
of these options also captures the positive traffic. The procedure
iteratively discards options from the candidate rule until it detects
that the rule is minimal, \ie{}, it finds that no other option can be
removed. The procedure uses negative traffic to guide the search.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }
\begin{algorithm}[t!]
  \noindent\hspace{-31ex}
  \textbf{INPUT:} overconstrained rule $or$ \\
  \noindent\hspace{-30ex} 
  \textbf{OUTPUT:} Set of minimal rules $\mathcal R$\\
%\vspace{-3ex}
\captionof{algorithm}{Minimization algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{or}\}$\Comment{initial population
    of candidate solutions contains only the input rule.}\label{line-init}
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{starting new
        iteration. new generation should be empty.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate individual/rule separately.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$      
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$
               \State $r[j]\gets 0$\Comment{discard option $j$ of rule
                 $r$, a copy of the original rule $\mathit{ro}$}
               \If{$\mathit{isnew(r)} \And \mathit{isfit}(r)$} 
                   \State $\mathcal R' \cup\gets \{r\}$\Comment{$r$
                     can lead to a plausible candidate solution.}\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets R'$\Comment{update new generation}
  \doWhile{$\mathcal R'\not= \{\}$}\Comment{repeat until reaching a fixpoint}\label{end-do-while}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode of the \tname{}
minimization algorithm. The algorithm takes as input the
overconstrained rule produced by the reverse engineering step and
produces a set $\mathcal R$ of rules that are minimal and only
captures the positive traffic. Line~\ref{line-init} initializes the
set $\mathcal{R}$ of candidate solutions with the input rule $or$. We
encode a rule as a bitvector where each index in the vector represents
one option; the value assigned at a given position indicates whether
or not the option is present in the rule. Each iteration of the outer
loop (lines~\ref{start-do-while}-\ref{end-do-while}) represents one
generation of the population of individuals produced by the
search. The algorithm terminates when no rule in the current
generation of individuals could be further reduced without violating
the property that a candidate solution should only capture positive
traffic. This means that execution has not reached
line~\ref{line-progress} as to augment the set $\mathcal{R'}$ and that
set will be empty at the end of one iteration of the outer
loop. Line~\ref{init-rprime} initializes the new generation of
candidate rules. The loop from
lines~\ref{start-foreach}--\ref{end-foreach} iterates through each
candidate rule in $\mathcal R$. Note that these rules have been found
in the previous generation. The algorithm mutates each of these rules
for $N$ times, discarding one option of the rule each time. The call
to $\mathit{isNew}$ checks if $r$ has not been previously generated
through another path whereas the call to $\mathit{isfit}$ checks if
the rule only captures positive traffic. Function $\mathit{copy}$
creates a copy of the bitvector of a rule, function $\mathit{len}$
returns the length of the vector, and function $\mathit{rand}$
randomly chooses one index of the vector.

\begin{theorem}
  The minimization algorithm computes rules that capture only
  the positive traffic.
\end{theorem}
\begin{proof}
  Every rule created during the search captures the positive traffic
  as every rule contains a proper subset of the option from the
  original rule and all those options are satisfied by the positive
  traffic. By definition of function $\mathit{isfit}$, these rules
  also capture \emph{only} the positive traffic. If some negative
  traffic was captured, a rule would not be added to $\mathcal
  R'$.\qed
\end{proof}

@@

The output of the minimization algorithm may contain rules that are
not minimal given the non-determinism of the algorithm. It is possible
that options that could be discarded are not selected in the loop from
lines~\ref{start-for}--\ref{end-for}. However, we found empirically
that a small value of $N$ suffices to produce minimal rules.

@@

\vspace{2ex}
\noindent\textbf{Limitation.} It is worth noting that although
\tname{} is accurate for generating rules that discrimate positive and
negative traffic, the quality of these rules is limited by the quality
of the training data it uses. For the positive traffic, \tname{} only
requires one sample, but can benefit of several samples. For the
negative traffic, \tname{} can use public datasets available online.

\vspace{2ex}
\noindent\textbf{Implementation details.} We implemented function
\emph{isFit} by 1)~updating Suricata to only use the informed rule,
2)~replaying the negative traffic (see
Section~\ref{sec:dataset-benign}), and 3)~reading the Suricata logs to
check if any benign message was captured by the rule. It is worth
noting that the benign traffic is stored in ``pcap'' files~\cite{pcap}
and Suricata can process those files very efficiently compared to
analyzing network messages directly.

@@

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,anonymous,pre-print]{acmart}

\pagenumbering{arabic} % Remove for camera-ready

@@

\vspace{1ex}
\noindent
\textbf{Differences.}~Note that there are differences between the
options produced in the first step of \tname{} and the options from
the golden rule, as per Figure~\ref{fig:synflood-example}. The option
\CodeIn{flags} does not mention parameters \CodeIn{1 2}, there is an
extra option \CodeIn{window}, and there are differences in the option
\CodeIn{threshold}. Considering the option \CodeIn{flags}, the first
parameter of that option indicates what flags in the packet must be
set while the second parameter indicates non-mandatory flags (\ie{},
``don't cares''). The option \CodeIn{flags: S} indicates that the flag
``S'' must be set and all flags not mentioned (``1'' and ``2'', in our
case) must be unset for the \nids\ to capture the
traffic. Consequently, that choice of option alone could result in
false negatives and false positives. An attack that satisfies all
other conditions in a rule and have flags ``1'' or ``2'' set would be
captured by the original rule but not the inferred rule---a case of
false negative. Likewise, an attack that satisfies all conditions and
have flags ``1'' and ``2'' unset would be captured by the inferred
rule but not the original rule---a case of false positive. The set of
positive examples is to blame in this case. To mitigate this issue,
the input messages should have covered cases with options ``1'' or
``2'' set. Considering option \CodeIn{windows: 64}, \tname{} is able
to generate rules without that option, as discussed above. Considering
option \CodeIn{threshold}, the values used in the golden rule are
domain-specific. \tname{} uses default values for these options. We
conjecture that the system administrator would be able to configure
this option.

@@

\begin{figure*}[ht!]
  \centering
  \includegraphics[trim=0 350 100 0,clip,width=0.65\textwidth]{figs/nids-workflow}
  \caption{\tname\ workflow.}
  \label{fig:overview}
  \vspace{-1ex}
\end{figure*}

@@

@@

The rule reported in the first step satisfies the criterion to only
capture positive traffic. Unfortunately, that rule is overconstrained.
The rule can include options that are irrelevant for the input
message. Consequently, the \nids\ may be unable to capture a different
manifestation of the same attack.

@@

\tname{} leverages public datasets~\cite{tcpreplay} of benign traffic
to guide the search for more general rules, hence addressing the
overffiting issue. To that end, \tname{} \emph{minimizes} the rules
produced in the previous step, preserving the invariant of \emph{only}
capturing the positive malicious traffic. We say that a rule is
minimal if discarding any option results in capturing negative (\ie{},
benign) traffic.  The minimization procedure we propose in this step
builds on the observation that all options from the rule produced in
the previous step are satisifed by the positive traffic. Consequently,
any subset of these options also captures the positive traffic. The
procedure iteratively discards options from the candidate rule until
it detects that the rule is minimal, \ie{}, it finds that no other
option can be removed. The procedure uses negative traffic to guide
the search.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }%
\begin{algorithm}[t!]
  \noindent\hspace{-25ex}
  \textbf{INPUT:} (overconstrained) seed rule $sr$ \\
  \noindent\hspace{-27.5ex} 
  \textbf{OUTPUT:} Set of minimized rules $\mathcal O$\\
%\vspace{-3ex}
\captionof{algorithm}{Minimization algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{sr}\}$\Comment{initial population
    of candidate solutions contains only the input
    rule.}\label{line-init}
  \State $\mathcal O \gets \emptyset$
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{new iteration. reset.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate rule.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$      
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$
               \State $r[j]\gets 0$\Comment{discard option $j$ in $r$}
               \If{$\mathit{isNew(r)} \And \mathit{isPlausible}(r)$} 
                   \State $\mathcal R' \gets \mathcal R' \cup \{r\}$\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets \mathcal R'$\Comment{update new generation}
       \State $\mathcal O \gets \mathcal O \cup \mathcal
       R$\Comment{remember all rules}
  \doWhile{$\mathcal R\not= \{\}$}\Comment{repeat until reaching a fixpoint}\label{end-do-while}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode of the \tname{}
minimization algorithm. The algorithm takes as input the
overconstrained rule produced by the reverse engineering step and
produces a set $\mathcal R$ of minimized rules. Line~\ref{line-init}
initializes the set $\mathcal{R}$ of candidate solutions with the seed
rule $sr$. We encode a rule as a bitvector where each index in the
vector represents one option; the value assigned at a given position
indicates whether or not the option is present in the rule. Each
iteration of the outer loop
(lines~\ref{start-do-while}-\ref{end-do-while}) represents one
generation of the population of individuals produced by the
search. The algorithm terminates when no rule in the current
generation of individuals could be further reduced without violating
the property that a candidate solution should only capture positive
traffic. This means that execution of one iteration has not reached
the statement declared at line~\ref{line-progress}, which adds an
element to the set $\mathcal{R'}$. Consequently, that set will be
empty at the end of one iteration of the outer
loop. Line~\ref{init-rprime} initializes the new generation of
candidate rules. The inner loop
(lines~\ref{start-foreach}--\ref{end-foreach}) iterates through each
candidate rule in $\mathcal R$. Note that these rules have been found
in the previous generation. The algorithm mutates each of these rules
for $N$ times, discarding one option of the rule each time. The call
to $\mathit{isNew}$ checks if $r$ has not been previously generated
through another path whereas the call to $\mathit{isPlausible}$ checks
if the rule only captures positive traffic. Function $\mathit{copy}$
creates a copy of the bitvector of a rule, function $\mathit{len}$
returns the length of the vector, and function $\mathit{rand}$
randomly chooses one index of the vector.

Note that no rules are produced in the last iteration of the
algorithm. Minimal rules are produced in the second-to-last
iteration. In the other iterations, intermediate rules are
produced. Conceptually, the search induces a search tree with the seed
rule $sr$ as the root, minimal rules as child nodes, and intermediate
rules as internal nodes. The edges of the tree denote the ancestor
relation between two rules. Note also that the algorithm reports every
rule in the tree instead of only minimal rules. The reason is that the
correct rule may be non-minimal. In summary, this algorithm reports
many rules that are plausible by using benign traffic to minimize an
overconstrained rule that is focused on the malicious traffic. In the
following step, \tname{} ranks the rules this algorithm reports by
their similarity with previous rules.

\begin{theorem}
  The minimization algorithm computes rules that capture only
  the positive traffic.
\end{theorem}
\begin{proof}
  Every rule created during the search captures the positive traffic
  as every rule contains a proper subset of the option from the
  original rule and all those options are satisfied by the positive
  traffic. By definition of function $\mathit{isfit}$, these rules
  also capture \emph{only} the positive traffic. If some negative
  traffic was captured, a rule would not be added to $\mathcal
  R'$.
\end{proof}

@@

\begin{proof}
The output of the minimization algorithm may contain rules that are
not minimal given the non-determinism of the algorithm. In principle,
it is possible that options that could be discarded are not selected
in the loop from lines~\ref{start-for}--\ref{end-for}. However, we
found empirically that a small value of $N$ suffices to discard those
options from a rule.
\end{proof}

@@

\vspace{2ex}
\noindent\textbf{Limitation.} It is worth noting that although
\tname{} is accurate for generating rules that discrimate positive and
negative traffic, the quality of these rules is limited by the quality
of the training data it uses. For the positive traffic, \tname{} only
requires one sample, but can benefit of several samples as the example
from Section~\ref{sec:dos} illustrated. The same applies for the
negative traffic. In this case, however, \tname{} can use public
datasets available online.

\vspace{2ex}
\noindent\textbf{Implementation details.} We implemented function
\emph{isPlausible} by 1)~updating Suricata to only use the rule that
is passed as parameter to this function, 2)~replaying the negative
traffic (see Section~\ref{sec:dataset-benign}), and 3)~reading the
Suricata logs to check if any benign message was captured by the
rule. It is worth noting that the benign traffic is stored in ``pcap''
files~\cite{pcap} and Suricata can process those files very
efficiently compared to analyzing network messages directly.

@@

\Mar{------------------------------------ estou aqui}
